---
title: "Week 4 Assignment"
author: "Harris Dupre"
date: "November 29, 2015"
output: html_document
---

We read in the dataset with wage information.

```{r}
data_wage = read.table(file="/Users/harrisdupre/Downloads/CPS08/cps08.csv",sep=",",header=TRUE)

summary(data_wage)
```

We can see that data_wage holds information on average hourly earnings, year, whether or not the observation is a bachelor, whether or not the observation is female, and the age.

Next we'll plot a histogram.

```{r}
hist(data_wage$ahe)
```

This shows us the frequency of different wages, between $10/hr and $20/hr being the most frequent wage.Next we'll create a boxplot that will compare the average hourly earnings of those who are single (denoted as 0), versus those who are married (denoted as 1).

```{r}
boxplot(ahe~bachelor,data=data_wage,main="Average Hourly Earnings vs Marital Status",xlab="Single/Married",ylab="AHE")
```

The average hourly earnings of those who are married have higher quartiles across the board, as well as a higher maximum, while maintaining the same minimum.

Finally we'll plot age against average hourly earnings.

```{r}
plot(data_wage$age,data_wage$ahe)
```

We can see that as age increases, maximum average hourly earnings tend to rise as well, although the minimum values persist regardless of age.